---
title: 
author: "Anna Moy"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### Data Preparation

```{r data-preparation, message = FALSE}
# load data
library(tidyverse)
library(psych)
library(infer)
library(openintro)
library(ggplot2)
```
```{r load-data, message = FALSE}
# read the csv file
salary <- read_csv("/Users/zhianna/Desktop/School/Data606/Salary_Data.csv")

dim(salary)
```
```{r tidy up}
# Tidying up the data
# renaming the column names
colnames(salary) <- c("age", "gender", "education", "job_title", "yrs_experience", "salary")

# remove missing data
salary <- salary %>%
  drop_na()

#add grouping years of experience
salary <- salary %>%
  mutate(yrs = case_when(yrs_experience <=  5 ~"0-5",
                         yrs_experience >  5 & yrs_experience <= 10 ~ "6-10",
                         yrs_experience > 10 & yrs_experience <= 15 ~ "11-15",
                         yrs_experience > 15 & yrs_experience <= 20 ~ "16-20",
                         yrs_experience > 20 ~ "20+"))

# order for the charts
salary <- salary %>%
  mutate(yrs = fct_relevel(yrs,"0-5","6-10","11-15","16-20","20+"))

# grouping for education into buckets
salary$degree<-ifelse(grepl("Bachelor's*", salary$education, ignore.case= TRUE), "Bachelor",
                ifelse(grepl("Master's*", salary$education, ignore.case= TRUE), "Master",
                ifelse(grepl("phd", salary$education, ignore.case= TRUE), "PhD",      
                ifelse(grepl("*High School", salary$education, ignore.case= TRUE),"High School","other"))))

#order for the charts
salary <- salary %>%
  mutate(degree = fct_relevel(degree, "High School", "Bachelor", "Master", "PhD"))

salary <- salary %>%
  mutate(range = case_when(age <=  29 ~"20-29",
                         age >  29 & yrs_experience <= 39 ~ "30-39",
                         age > 39 & yrs_experience <= 49 ~ "40-49",
                         age > 49 & yrs_experience <= 59 ~ "50-59",
                         age >59  ~ "60+"))
# order for the charts
salary <- salary %>%
  mutate(range = fct_relevel(yrs,"20-29","30-39","40-49","50-59","60+"))

salary

```

### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Is there a relationship between salary and gender, years of experience, age and education?

### Cases 

**What are the cases, and how many are there?**

Each case represents a worker's demographic data and their employment information .  There are 6704 observations in the dataset.

### Data collection 

**Describe the method of data collection.**

The data was collected through different industries and regions across the globe. It was sourced through employment websites and surveys. The data contains information about age, gender, education, job titles, years of experience, salary, race and geographic location. 
 

### Type of study 

**What type of study is this (observational/experiment)?**

This information was received through observational. 

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

Data was collected by Kaggle: [Salary by Job Title and Country](https://www.kaggle.com/datasets/amirmahdiabbootalebi/salary-by-job-title-and-country).

### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is salary which is quantitative and numeric.

### Independent Variable(s)

The explanatory variable are :
  
  - Gender   (qualitative and categorical)
  
  - Education (qualitative and categorical)
  
  - Years of experience (quantitative and numeric)
  
  - Age (quantitative and numeric)

### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r summary-statistics}
##filter for only female and male (ignoring Others)
gender <- salary %>%
  filter(gender == "Female" | gender == "Male")
gender
```
Salary for Female and Male
```{r salary-female-and-male}
#observe the salary difference in mean between male and female 
diff_mean <- gender %>%
  group_by(gender) %>%
  summarize(mean_salary = mean(salary)) %>%
  pull() %>%
  diff()
diff_mean

#average salary for female and male only
avg <- gender %>%
  summarize(mean = mean(salary))
avg


# filter for female only
female <- gender %>%
  filter(gender == "Female") %>%
  select(salary)

# SD and mean for female
sdfm <- sd(gender$salary)
meanfm <- mean(gender$salary)

# plot to see if it looks like a normal distribution
ggplot(data = female, aes(x = salary)) + geom_blank() +
  geom_histogram(aes(y = ..density..)) +
  stat_function(fun = dnorm, args = c(mean = meanfm, sd = sdfm), col = "tomato") +
  xlab("Female Salaries")

#filter for male only
male <- gender %>%
  filter(gender == "Male")

#SD and mean for male
sdm <- sd(gender$salary)
meanm <- mean(gender$salary)

#Plot to see if it looks like a normal distribution
ggplot(data = male, aes(x = salary)) + geom_blank() +
  geom_histogram(aes(y = ..density..)) +
  stat_function(fun = dnorm, args = c(mean = meanm, sd = sdm), col = "tomato") + 
  xlab("Male salaries")

#qqplot for female and male
qqnormsim(sample = salary, data = female)
qqnormsim(sample = salary, data = male)
```
```{r boxplot}
# avg of gender
box <- gender %>%
  group_by(gender) %>%
  summarize(mean = round(mean(salary)))
box

# boxplot data
text <- gender %>%
  group_by(gender) %>%
  summarize(median = round(median(salary)), mean = round(mean(salary)), min = min(salary), max = max(salary), q1 = quantile(salary, .25), q3 = quantile(salary, .75), iqr = q3-q1)
text

text$labs = c()

#boxplot
ggplot(gender, aes(gender, salary)) +
 geom_boxplot() +
 geom_point(data = box, aes(gender, y = mean),color ="blue", size = 4) +
 geom_text(data = text, aes(gender, y = median, label = median, vjust = -2)) +
 geom_text(data = text, aes(gender, y = max, label = max, vjust = -.5)) +
 geom_text(data = text, aes(gender, y = min, label = min, vjust = -.1)) +
 geom_text(data = text, aes(gender, y = q1, label = q1, vjust = -.3)) +
 geom_text(data = text, aes(gender, y = mean, label = mean, vjust = -.3)) +
 geom_text(data = text, aes(gender, y = q3, label = q3, vjust = -.5)) 
 

#density plot
ggplot(gender, aes(x=salary, color = gender)) + 
  geom_density()
#Filter for female
sum <- gender %>%
  filter(gender == "Female") %>%
  summarize(total = median(salary))
sum

#quartile for 25% and 75%
q1 <- quantile(sum$total, .25)
q3 <- quantile(sum$total, .75)
iqr <- q3 - q1


# filter for male
sum2 <- gender %>%
  filter(gender == "Male") %>%
  summarize(total = median(salary))
sum2

#quartile for 25% and 75%
q12 <- quantile(sum$total, .25)
q32 <- quantile(sum$total, .75)
iqr2 <- q3 - q1


```
Years of Experience and Salary Analysis
```{r more analysis}
#group years of experience
gender %>%
  group_by(yrs) %>%
  summarize( x_bar =mean(salary),sd =sd(salary),n=n())

#plot the data points
ggplot(gender, aes(x = salary, y = yrs_experience)) +
  geom_point() +
  geom_smooth(method ="lm", formula= y ~x, color = "darkgreen", se = FALSE)

#bar chart for salary and years of experience
ggplot(gender, aes(yrs, salary)) +
  geom_bar(stat = "identity")

# correlation for salary and experience
cor(gender$salary, gender$yrs_experience)

ex_lm <- lm(salary~yrs_experience, data = gender)
summary(ex_lm)

```

Education and Salary Analysis
```{r education}
gender %>%
  group_by(degree) %>%
  summarize( x_bar =mean(salary),sd =sd(salary),n=n())

# Mean for each degree
deg_mean <- gender %>%
  group_by(degree) %>%
  summarize(mean=mean(salary))
deg_mean

# plot boxplot for degree and salary with mean
ggplot(gender, aes(degree, salary)) +
  geom_boxplot() +
  geom_point(data = deg_mean, aes(degree, y = mean), color ="blue", size = 2)


#Anova for degree and salary
degree_aov <- aov(salary ~ degree, data = gender)
degree_aov
summary(degree_aov)
```

Age and Salary Analysis
```{r age}
gender %>%
  summarize( x_bar =mean(salary),sd =sd(salary),n=n())

# average for each age in salary
age_mean <- gender %>%
  summarize(mean=mean(salary))
age_mean

# scatter plot for age and salary
ggplot(gender, aes(x=age, salary)) + 
  geom_point() +
  geom_smooth(method ="lm", formula= y ~x, color = "red", se = FALSE)

# correlation for salary and age
cor(gender$salary, gender$age)

age_lm <- lm(salary ~ age, data = gender)
summary(age_lm)

```
```{r plot}
library(ggplot2)
library(GGally)

plot <- gender %>%
  select(salary, gender, yrs_experience, education)

ggpairs(plot)
```

```{r multiple-regression}
lm_out <- lm(salary ~ gender + yrs_experience + degree, data = gender)
summary(lm_out)

```
